---
title: "STAT 345 Midterm Project"
font: 12pt
date: "April 3rd, 2026"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

> "Our offense is like the Pythagorean Theorem. There is no answer!", *Shaquille O'Neal*

## REPORT:

```{r, echo = FALSE, message = FALSE}
# add libraries
library(tidyverse)
library(hoopR)
library(dplyr)
library(ggplot2)
library(knitr)
library(ggpubr)
library(png)
library(gganimate)
library(gifski)

# load in data from hoopR for seasons 2015 to 2026
data <- hoopR::load_nba_pbp(seasons = 2015:2026)
```

```{r, echo = FALSE, message = FALSE}
# filters data to the New Orleans Pelicans (team 3) and only shooting plays
# adds new column to classify makes/misses
# gets rid of unnecessary columns so data is easier to manage and loads faster
baby_data <- data |> 
  filter(team_id == 3, shooting_play == TRUE) |> 
  mutate(make_miss = ifelse(score_value > 0, "Make", "Miss")) |> 
  select("score_value", "team_id", "shooting_play", "coordinate_x_raw", "coordinate_y_raw", "season", "make_miss")
```

```{r, echo = FALSE, message = FALSE}
# gets background image of court for animation
img <- readPNG("midterm_bbcourt.png")

# creates a plot based on raw x and y coords
# adds background image of NBA court
# adjust opacity (alpha) so darker AND added weight/height so dots are larger for multiple shots from same coords
# assign colors for made/missed shots...made shots = Pelicans Navy, missed = Pelicans Red
# added labels for title and legend
# fix coords for more accurate alignment with background image
# add animation between seasons, adjusting transition/state length for better viewing
# removes numbers and ticks from axes
ggplot(baby_data, aes(x = coordinate_x_raw, y = coordinate_y_raw)) + 
  background_image(img) + 
  geom_jitter(aes(color = make_miss), width = 0.5, height = 0.5, alpha = 0.1, size = 1) + 
  scale_color_manual(values = c("Make" = "#0C2340", "Miss" = "#C8102E")) + 
  labs(title = "New Orleans Pelicans Shots {closest_state}", x = "", y = "", color = "Shot Result") + 
  coord_fixed(xlim = c(0, 50), ylim = c(0, 36.75)) + 
  transition_states(season, transition_length = 3, state_length = 100) +
  theme(axis.ticks = element_blank(), axis.text = element_blank())
```

### **Summary:**

-   **Most Common Shot Locations:** under the basket, free throw line, corners, and top of three point line

-   **Shift Across Seasons:** shot distribution slowly shifts from a wide spread to a condensed shot selection of just a few main spots

-   **Make/Miss Ratio:** made shots appear more in the common shot locations and increase over time

-   **Suggestions for Improvement:** focus on improving shot performance or shot location to increase overall scoring success

### **Explanation:**

-   **Most Common Shot Locations:**

    -   The majority of shot attempts are made directly under the basket, on the free throw line, or along the three point line. On the three point line, there is a preference for the corners and wings, with fewer in between. The highest concentration of shot attempts are under the basket. This aligns with the common strategy for most basketball teams, which is to focus on point efficiency by prioritizing either close range shots and layups for easy points or three point shots for more points.

-   **Shift Across Seasons:**

    -   Looking at all the seasons, from 2015 to 2026, shot distribution gradually shifts. In the early seasons, there is a large spread of shots all over the court. Then, over time, the shot distribution narrows down to most being taken under the basket, on the free throw line, in the corners, or on the wings and less being taken in the mid range area. This indicates a clear improvement in playing style to prioritize scoring efficiency.

-   **Make/Miss Ratio:**

    -   The majority of made shots tend to be concentrated under the basket, where scoring is easier and more consistent, as well as on the free throw line in later seasons, where shots are uncontested. Over time, the data shows improvement in shot successes. In the early seasons you can clearly see there are more red dots, especially for free throws. It slowly shifts from red to blue in the later seasons , showing an increase in made free throws. There also appears to be a greater percentage of made shots beneath the basket in the later seasons compared to the early seasons.

-   **Suggestions for Improvement:**

    -   To improve overall scoring success, I suggest focusing on shot performance and better ball movement. Quality shots and opening up players would help increase the percentage of made shots. Another suggestion is to focus on taking more close range shots rather than three-pointers, since there is already higher success under the basket, which could lead to higher scoring output.

## ABOVE AND BEYOND:

### Analysis/Reasoning:

The NBA three point line measures at 23.75 feet straight out from the base line. According to <https://www.statmuse.com/nba/ask/pelicans-3-point-shooting-percentage>, the Pelican's shooting percentage for three pointers in 2025-2026 is 34.7%. Using that percentage, an expected value for threes would be 3 \* 34.7% = 1.04. Based on that, the average make percentage for an imagined 4 point shot would have to be 1.04 / 4 = 26.03%. Then, to determine where to put the new 4 point line, I calculated how far out the Pelicans were making roughly 26.03% of their shots.

```{r}
part4data <- baby_data |> filter(coordinate_y_raw >= 28)
makeCount <- sum(part4data$make_miss == "Make")
percent28 <- makeCount / nrow(part4data)
percent28

part4data <- baby_data |> filter(coordinate_y_raw >= 27)
makeCount <- sum(part4data$make_miss == "Make")
percent27 <- makeCount / nrow(part4data)
percent27
```

Based on the above code, I've determined the distance for a 4 point line should be 27.5 feet out from the base line. The Pelicans make roughly 28.59% of their shots at or beyond 27 feet from the base line, but only 22.96% of their shots at or beyond 28 feet. This makes 27.5 feet a happy medium that should get the make percentage close to 26%. Doing some simple math, 27.5 - 23.75 = 3.75, the 4 point line would follow the current 3 point line, but be 3.75 feet further from the basket.

### Consequences:

The addition of a 4 point line at 27.5 feet from the base line would have positive results for teams with players who consistently score deep threes. For anyone who can consistently score beyond 27.5 feet, all those former three point shots become four point shots, increasing their overall points per game which in turn increases their team's overall scoring output, causing their team to be more successful in games.

Assuming my favorite player on the New Orleans Pelicans is Trey Murphy III, who currently has a 38.3 three point percentage, and, according to <https://thebirdwrites.com/triggad-trey-murphy-iiis-elite-shooting-sparks-pelicans/>, 41.8% of his threes are from 25-29 feet. With that kind of range, he would definitely be making a significant amount of 4 pointers, increasing his average points per game and giving his team a huge offensive advantage.
